---
title: "The effects of body tissue and gene drift on the differential gene expression of Aphedius ervi"
author: "Nanda Vo"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    theme: cerulean
  pdf_document:
    toc: yes
---
GLMM:

https://myles-lewis.github.io/glmmSeq/articles/glmmSeq.html#citing-glmmseq
https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

PCA:
https://knowledge.dataiku.com/latest/ml-analytics/statistics/concept-principal-component-analysis-pca.html
https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#dispersion-plot-and-fitting-alternatives


# Introduction

This markdown file serves as workflow for the differential gene expression analysis on Aphedius ervi.
In this thesis, I study two families of A. ervi, that were separately inbreded beforehand, to achieve genetic homogenity.

The dataset consists of three treatment compartemnts: Line, Conditioned, and Experience.
Line describes the lineage of the subjects. All parental generations were exclusively reared on the same type of host, PA (pea aphid) or FA (foxglove aphid).
Conditioned describes the type of host, where the subjects emerged from.
Experience describes the type of host, that was forced on the subject for ovipositioning. 

In this thesis, I aim to study the relative contributions of the treatments (treatment components) on the differential gene expression of the studied individuals.


```{r eval=FALSE, include=FALSE}
# Install CRAN packages
install.packages(c(
  "car", "ggplot2", "ggpubr", "lme4", "lmerTest", "qvalue",
  "knitr", "kableExtra", "tidyverse", "plyr", "dplyr","pvclust"
))

# Install Bioconductor packages
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install(c("glmmSeq", "limma", "edgeR", "DESeq2"))

```


## Load needed libraries


```{r Libraries, echo=TRUE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(pvclust)
library(glmmSeq)
library(MASS)
library(car)
library(ggplot2)
library(ggpubr)
library(lme4)
library(lmerTest)
library(qvalue)
library(limma)
library(edgeR)
library(knitr)
library(kableExtra)
library(qvalue)
library(caret)
library(lmerTest)
library(glmmSeq)
library(viridis)
library(DESeq2)
library(plyr)
library(dplyr)
library(DESeq2)
library(edgeR)
library(pheatmap)
library(RColorBrewer)
library(parallel)
library(ggpubr)  # For adding p-values to plots
library(ComplexHeatmap)
library(circlize)
library(kableExtra)
library(VennDiagram)

# Set knitr root directory for knitting
knitr::opts_knit$set(root.dir = "C:/Users/nanda/Desktop/Master Thesis")
```


## Load the pre-processed count and sample tables into R.


```{r Load data, echo=TRUE}
setwd("C:/Users/nanda/Desktop/Master Thesis")
# Load count data
counts_data <- read.csv("counts.csv")
# Set gene column as row names
rownames(counts_data) <- counts_data$Gene
# Remove Gene as a column
counts_data$Gene <- NULL

# Load sample data
samples_data <- read.csv("samples.csv")
# Set Sample column as row names
rownames(samples_data) <- samples_data$Sample
# Remove Sample as a column
samples_data$Sample <- NULL
# Normalize Treatment column
samples_data$Treatment <- gsub("-", "_", samples_data$Treatment)

# Filter out samples where 'Line' contains "G(FA)"
samples_data <- samples_data %>%
  filter(!grepl("G\\(FA\\)", Line))

# Ensure counts_data has columns that match filtered samples_data rows
counts_data <- counts_data[, colnames(counts_data) %in% rownames(samples_data)]

# Double check matching rows and columns of samples and counts
if (!all(colnames(counts_data) %in% rownames(samples_data))) {
  stop("Mismatch between samples and counts data.")
}
if (!all(colnames(counts_data) == rownames(samples_data))) {
  stop("Order of samples and counts data do not match.")
}
```


## Filter data by CPM
Now, we filter or count data by removing genes with a CPM (counts per million) below 1, using edgeR.
Out of 20226 genes, we removed 1881.

```{r Filter by CPM, include=TRUE}
# Create a DGEList object for edgeR
dge <- DGEList(counts = counts_data)

# Calculate CPM
cpm_data <- cpm(dge)

# Define a filter threshold (1 CPM)
cpm_threshold <- 1
keep <- rowSums(cpm_data >= cpm_threshold) >= 1
dge <- dge[keep,]
dim(dge)

# Filter the count data
filtered_counts_data <- counts_data[keep, ]
```


## DESeq2 dds object
DGE analysis with DESeq2 using following model: counts ~ Line + Conditioned + Experience + Tissue + Family.
I use DESeq2 to get an overall impression of the differentiality of gene expression profiles. This means that I aim to compare the overall expression profile of the individuals for the beginning.



```{r DGE model, echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE, collapse=TRUE}
# Create DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = filtered_counts_data,
                              colData = samples_data,
                              design = ~  Line + Conditioned + Experience + Tissue + Family)

# Perform differential expression analysis
dds <- DESeq(dds)

#save dds object
#saveRDS(dds, file = "dds_model.rds")
  #load it
  #dds <- readRDS("dds_model.rds")


```


## Gene dispersion and normalized counts
Calculate dispersions and normalize counts.
The dispersion represents the variability of gene expression levels.

The normalization by size factors corrects for sequencing depths
(important for avoiding false negatives for lowly expressed genes)

Also apply variance stabilizing transformation on data



```{r Dispersion and count normalization, echo=TRUE, fig.show='hold', collapse=TRUE}

 # Estimate dispersions (this is already done during the DESeq call)
        dds <- estimateDispersions(dds)
        dispersions <- dispersions(dds)
        dispersions <- setNames(dispersions, rownames(filtered_counts_data))
        
        # Estimate size factors (this is also done during DESeq normalization)
        dds <- estimateSizeFactors(dds)
        size_factors <- sizeFactors(dds)

        # Normalize counts by size factors
        normalized_counts <- counts(dds, normalized = TRUE)
      
      
        # Plot dispersion estimates
        plotDispEsts(dds)

      
        # Variance stabilizing transformation (VST); DESeq2 usually uses normalized counts for the VST 
        vst_dds <- vst(dds)
        
        # You can use the transformed data for PCA or clustering
        vst_counts <- assay(vst_dds)
```


## PCA by tissue
Do PCA with the top 500 most variable genes. For the beginning, I'm separating by tissue type.



```{r PCA by tissue type, echo=TRUE, fig.show='hold', collapse=TRUE}
# PCA plot by tissue
# Identify the variance for each gene
gene_variances <- apply(vst_counts, 1, var)

# Select the top 500 most variable genes
top_500_genes <- order(gene_variances, decreasing = TRUE)[1:500]

# Subset the VST data to include only these top 500 genes
vst_counts_top_500 <- vst_counts[top_500_genes, ]

# Perform PCA on the filtered VST data for the top 500 genes
pca_top_500 <- prcomp(t(vst_counts_top_500), scale. = TRUE)

# Calculate the percentage of variance explained by PC1 and PC2
explained_variance <- summary(pca_top_500)$importance[2, ]
pc1_var <- round(explained_variance[1] * 100, 2)
pc2_var <- round(explained_variance[2] * 100, 2)

# Create a data frame with PCA results and sample metadata
pca_data_top_500 <- as.data.frame(pca_top_500$x)
pca_data_top_500$Tissue <- samples_data$Tissue

# Enhanced PCA plot with borders only on x and y axes
library(ggplot2)

# Create PCA plot with publication-ready customizations
ggplot(pca_data_top_500, aes(x = PC1, y = PC2, color = Tissue)) +
  geom_point(size = 3, alpha = 0.9) +  # Increase point opacity for distinct visualization
  scale_color_brewer(palette = "Dark2", name = "Tissue origin",  # Use a color palette similar to the reference
                     labels = c("Head", "Ovaries", "Venom glands")) + 
  labs(
    title = "PCA of All Tissue Origins (Top 500 Most Variable Genes)",
    x = paste0("PC1 (", pc1_var, "% variance)"),
    y = paste0("PC2 (", pc2_var, "% variance)")
  ) +
  theme_minimal(base_size = 14) +  # Larger base font size for readability
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title and make it bold
    axis.title = element_text(face = "bold"),  # Bold axis titles
    legend.position = "bottom",  # Place legend at the bottom
    legend.key = element_blank(),  # Remove legend key rectangles
    panel.grid.major = element_line(color = "gray80"),  # Subtle major grid lines
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    axis.line = element_line(color = "black", linewidth = 0.8)  # Add borders only to x and y axes
  )
        
```

## PCA within tissue by family
Since the expression profiles are clearly clustered by tissue (and high percentages for PC 1), 
I'll do a PCA within each tissue type, separating by family as well.
The separation implies that each tissue type has a distinct group of expressed genes. I'll also verify it later by creating tables showing the number of overlapping genes between tissue types. (glmmSeq section)


```{r PCA within tissue, separating by family, echo=TRUE, fig.show='hold', collapse=TRUE}
# Load necessary libraries
library(ggplot2)
library(gridExtra)
library(grid)
library(viridis)

# Custom publishable theme for consistent styling
theme_publishable <- function() {
  theme_minimal() +
    theme(
      panel.grid = element_blank(),  # Remove grid lines
      axis.line = element_line(color = "black", size = 1),  # Thick axis lines
      axis.ticks = element_line(color = "black", size = 0.8),  # Thick axis ticks
      legend.position = "none",  # Remove individual plot legends
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),  # Center and bold title
      axis.title = element_text(face = "bold", size = 12),  # Bold axis titles
      axis.text = element_text(size = 10)  # Adjust axis text size
    )
}

# Function to extract the legend as a "shared x-axis"
get_shared_legend <- function() {
  legend_plot <- ggplot(data.frame(x = c(1, 2), y = c(1, 2), label = c("F34", "F39")), aes(x, y, color = label)) +
    geom_point(size = 4) +
    scale_color_manual(
      values = c("F34" = "#8856a7", "F39" = "#f03b20"),  # Purple and Orange
      name = "Family"
    ) +
    guides(color = guide_legend(override.aes = list(size = 5))) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      legend.title = element_text(face = "bold", size = 12),
      legend.text = element_text(size = 10),
      panel.grid = element_blank(),
      axis.line = element_blank(),
      axis.ticks = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
  
  # Extract legend as grob
  return(get_legend(legend_plot))
}

# Function to perform PCA and create styled plots
perform_pca <- function(tissue_name, normalized_counts, samples_data) {
  
  # Subset samples_data for the specified tissue
  tissue_samples <- samples_data[samples_data$Tissue == tissue_name, ]
  
  # Subset normalized counts data for the specified tissue
  tissue_counts <- normalized_counts[, rownames(tissue_samples)]
  
  # Identify the variance for each gene
  gene_variances <- apply(tissue_counts, 1, var)
  
  # Select the top 500 most variable genes
  top_500_genes <- order(gene_variances, decreasing = TRUE)[1:500]
  
  # Subset the data to include only the top 500 most variable genes
  tissue_counts_top_500 <- tissue_counts[top_500_genes, ]
  
  # Remove zero-variance columns
  zero_variance_filter <- apply(tissue_counts_top_500, 1, var) > 0
  tissue_counts_top_500 <- tissue_counts_top_500[zero_variance_filter, ]
  
  # Perform PCA
  pca_result <- prcomp(t(tissue_counts_top_500), scale. = TRUE)
  
  # Calculate percentage of variance explained by PC1 and PC2
  explained_variance <- summary(pca_result)$importance[2, ]
  pc1_var <- round(explained_variance[1] * 100, 2)
  pc2_var <- round(explained_variance[2] * 100, 2)
  
  # Create a data frame with PCA results and metadata
  pca_data <- as.data.frame(pca_result$x)
  pca_data$Family <- tissue_samples$Family
  
  # Adjust the title to use tissue names
  plot_title <- switch(
    tissue_name,
    "head" = "Head",
    "ovaries" = "Ovaries",
    "venom" = "Venom Glands"
  )
  
  # Create PCA plot with Purple and Orange colors
  pca_plot <- ggplot(pca_data, aes(x = PC1, y = PC2, color = Family)) +
    geom_point(size = 3, alpha = 0.9) +  # Adjusted point size and transparency
    scale_color_manual(
      values = c("F34" = "#8856a7", "F39" = "#f03b20"),  # Purple and Orange
      name = "Family"
    ) +
    labs(
      title = plot_title,
      x = paste0("PC1 (", pc1_var, "% variance)"),
      y = paste0("PC2 (", pc2_var, "% variance)")
    ) +
    theme_publishable()
  
  return(pca_plot)
}

# Create individual plots for each tissue
pca_head_plot <- perform_pca("head", normalized_counts, samples_data)
pca_ovaries_plot <- perform_pca("ovaries", normalized_counts, samples_data)
pca_venom_plot <- perform_pca("venom", normalized_counts, samples_data)

# Get shared legend as a "shared x-axis"
shared_legend <- get_shared_legend()

# Combine the plots with the shared legend
final_plot <- grid.arrange(
  arrangeGrob(
    pca_head_plot,
    pca_ovaries_plot,
    pca_venom_plot,
    ncol = 3
  ),
  shared_legend,
  ncol = 1,
  heights = c(10, 1)  # Shared legend acts as a bottom "x-axis"
)

# Save the final plot as a high-resolution TIFF
ggsave("PCA_Final_Plots_with_Shared_X_Legend.tiff", plot = final_plot, width = 16, height = 6, dpi = 300, units = "in")

```
Head and venom tissues show clear separation between families, mostly represented by PC 1.
This implies that the family has the biggest effect on the separation between those two distinct clusters.
For the ovaries, the families show two distinct clusters as well, but mostly represented by PC 2. Consequently, there may be another factor that contributes more in the variance of the mean ovary expression profiles (of the top 500 most variable genes) of the samples.




```{r Loading bar plots, separating by family, echo=TRUE, fig.show='hold', collapse=TRUE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(caret)
library(gridExtra)
library(viridis)
library(grid)

# Custom publishable theme
theme_publishable <- function() {
  theme_minimal() +
    theme(
      panel.grid = element_blank(),  # Remove grid lines
      panel.border = element_blank(),  # Remove full plot border
      axis.line = element_line(color = "black", size = 1),  # Add lines only to the axes
      axis.ticks = element_line(color = "black"),  # Add axis ticks
      axis.text.x = element_blank(),  # Remove x-axis text labels
      axis.ticks.x = element_blank(),  # Remove x-axis ticks
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      legend.box = "horizontal",
      legend.text = element_text(size = 10),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.title.x = element_blank(),  # Remove x-axis title
      axis.title.y = element_text(face = "bold", size = 14)  # Enable y-axis title for tissue labels
    )
}

# Function to create loading plots for specified PCs
create_loading_plots <- function(pcs = c(1, 2, 3, 4)) {
  all_plots_list <- list()
  tissue_labels <- c("Head", "Ovaries", "Venom Glands")  # Labels for the rows
  
  row_counter <- 1  # Track rows for assigning tissue labels
  
  for (tissue in unique(samples_data$Tissue)) {
    cat("\nPerforming PCA and generating loading plots for Tissue:", tissue, "\n")
    
    # Subset the samples_data and vst_counts for this tissue type
    tissue_samples_data <- samples_data[samples_data$Tissue == tissue, c("Line", "Conditioned", "Experience", "Family")]
    tissue_vst_counts <- vst_counts[, samples_data$Tissue == tissue]
    
    # Convert categorical variables to one-hot encoded numeric variables for the sub-treatments
    dummies <- dummyVars("~ .", data = tissue_samples_data)
    sub_treatments_numeric <- predict(dummies, newdata = tissue_samples_data)
    
    # Identify the variance for each gene for this tissue
    gene_variances <- apply(tissue_vst_counts, 1, var)
    
    # Select the top 500 most variable genes for this tissue
    top_500_genes <- order(gene_variances, decreasing = TRUE)[1:500]
    
    # Subset the VST data to include only the top 500 most variable genes
    vst_counts_top_500 <- tissue_vst_counts[top_500_genes, ]
    
    # Transpose the VST counts so that samples are rows and genes are columns
    vst_counts_top_500_t <- t(vst_counts_top_500)
    
    # Perform PCA on the combined data (samples as rows, sub-treatments and genes as variables)
    combined_data <- cbind(sub_treatments_numeric, vst_counts_top_500_t)
    pca_tissue <- prcomp(combined_data, scale. = TRUE)
    
    # Extract loadings (contributions) for the specified PCs
    loadings <- as.data.frame(pca_tissue$rotation[, pcs])
    loadings$variable <- rownames(loadings)
    
    # Filter the loadings for sub-treatments only
    sub_treatment_loadings <- loadings[rownames(loadings) %in% colnames(sub_treatments_numeric), ]
    
    # Combine loadings for each main variable (sum of absolute loadings across levels)
    sub_treatment_loadings$main_variable <- gsub("[0-9A-Z]+$", "", sub_treatment_loadings$variable)  # Properly label main variables
    loadings_summary <- sub_treatment_loadings %>%
      group_by(main_variable) %>%
      summarize(across(starts_with("PC"), ~ sum(abs(.)), .names = "abs_{.col}"))  # Sum absolute loadings
    
    # Order the variables for plotting
    variable_order <- c("Family", "Line", "Conditioned", "Experience")
    loadings_summary$main_variable <- factor(loadings_summary$main_variable, levels = variable_order)
    
    # Create a list to store the plots for specified PCs
    for (pc in pcs) {
      pc_column <- paste0("abs_PC", pc)
      
      # Dynamically create a data frame for plotting
      loadings_to_plot <- data.frame(
        main_variable = loadings_summary$main_variable,
        loading_value = loadings_summary[[pc_column]]
      )
      
      # Plot for each PC
      p <- ggplot(loadings_to_plot, aes(x = main_variable, y = loading_value, fill = main_variable)) +
        geom_bar(stat = "identity", color = "black") +  # Add border to bars
        scale_fill_viridis_d(option = "C", direction = 1, name = "Sub-Treatments") +
        labs(
          title = if (row_counter == 1) paste("Loadings for PC", pc) else NULL,
          y = if (pc == pcs[1]) tissue_labels[row_counter] else NULL  # Add tissue label as y-axis title for the first column
        ) +
        theme_publishable() +
        theme(
          legend.position = "none",  # Remove legends for individual plots
          axis.text.x = element_blank(),  # Remove x-axis labels
          axis.ticks.x = element_blank()  # Remove x-axis ticks
        )
      
      # Add the plot to the list
      all_plots_list <- c(all_plots_list, list(p))
    }
    
    # Increment the row counter for tissue labels
    row_counter <- row_counter + 1
  }
  
  # Combine all plots into a grid
  arranged_plots <- grid.arrange(
    grobs = all_plots_list,
    ncol = 4,
    nrow = 3,
    left = textGrob("Absolute Loadings", rot = 90, vjust = 0.5, gp = gpar(fontsize = 14, fontface = "bold")),  # Shared y-axis label
    bottom = get_legend(all_plots_list[[1]] + theme(legend.position = "bottom"))  # Shared legend
  )
  
  return(arranged_plots)
}

# Create and display the final plot
final_plot <- create_loading_plots(pcs = c(1, 2, 3, 4))

# Save the final plot
ggsave("final_plot_with_correct_tissue_labels.tiff", final_plot, width = 16, height = 12, dpi = 300, units = "in")


```


## Volcano plots within tissue by family
The following volcano plots underline these findings.


```{r Volcano plots within each tissue, separating by family, echo=TRUE, fig.show='hold', collapse=TRUE}
 #volcano plots between families within each treatment
          # Define the volcano plot function for comparing Family F34 vs F39 within a tissue
          perform_volcano_plot_family_comparison <- function(tissue_name, normalized_counts, samples_data) {
            
            # Subset the samples_data for the specified tissue
            tissue_samples <- samples_data[samples_data$Tissue == tissue_name & samples_data$Family %in% c("F34", "F39"), ]
            
            # Subset the normalized counts data for the specified tissue
            tissue_counts <- normalized_counts[, rownames(tissue_samples)]
            
            # Ensure the Family column is recognized as a factor
            tissue_samples$Family <- factor(tissue_samples$Family)
            
            # Calculate log2 fold changes and p-values between Family F34 and F39
            log2_fc <- apply(tissue_counts, 1, function(gene_counts) {
              mean(gene_counts[tissue_samples$Family == "F39"]) - 
                mean(gene_counts[tissue_samples$Family == "F34"])
            })
            
            # Perform t-tests to get p-values
            p_values <- apply(tissue_counts, 1, function(gene_counts) {
              t.test(gene_counts[tissue_samples$Family == "F34"], 
                     gene_counts[tissue_samples$Family == "F39"])$p.value
            })
            
            # Create a data frame with the results
            results <- data.frame(
              Gene = rownames(tissue_counts),
              log2FC = log2_fc,
              p_value = p_values,
              neg_log10_p_value = -log10(p_values)
            )
            
            # Filter out rows with NA, Inf, or -Inf values
            results <- results %>%
              filter(!is.na(log2FC) & !is.infinite(log2FC) &
                       !is.na(neg_log10_p_value) & !is.infinite(neg_log10_p_value))
            
            # Determine significance (adjust as needed, e.g., p < 0.05 and |log2FC| > 1)
            results$Significant <- "Not Significant"
            results$Significant[results$p_value < 0.05 & results$log2FC > 1] <- "Up"
            results$Significant[results$p_value < 0.05 & results$log2FC < -1] <- "Down"
            
            # Create the volcano plot comparing F34 vs F39
            volcano_plot <- ggplot(results, aes(x = log2FC, y = neg_log10_p_value, color = Significant)) +
              geom_point(size = 2) +
              scale_color_manual(values = c("Not Significant" = "gray", "Up" = "red", "Down" = "blue")) +
              labs(title = paste("Volcano Plot for", tissue_name, "Tissue - F34 vs F39"),
                   x = "Log2 Fold Change (F39 vs F34)",
                   y = "-Log10(P-Value)") +
              geom_vline(xintercept = c(-1, 1), linetype = "dashed") +  # Add vertical lines at log2FC = ±1
              geom_hline(yintercept = -log10(0.05), linetype = "dashed") +  # Add horizontal line at p = 0.05
              theme_minimal()
            
            return(volcano_plot)
          }
          
          # Example usage for all tissues
          volcano_head_plot <- perform_volcano_plot_family_comparison("head", normalized_counts, samples_data)
          volcano_ovaries_plot <- perform_volcano_plot_family_comparison("ovaries", normalized_counts, samples_data)
          volcano_venom_plot <- perform_volcano_plot_family_comparison("venom", normalized_counts, samples_data)
          
          # Display the plots for F34 vs F39 in each tissue
          print(volcano_head_plot)  # Volcano plot for head tissue (F34 vs F39)
          print(volcano_ovaries_plot)  # Volcano plot for ovaries tissue (F34 vs F39)
          print(volcano_venom_plot)  # Volcano plot for venom tissue (F34 vs F39)
```
The majority of the individuals genes show a clear up- (positive LFC) or downregulation (negative LFC).




## GLMMs for each tissue
### GLMM for head
Finally, I'll proceed with GLMMs to identify sets of significantly differentially expressed genes.
Due to the high differentiality caused by the families (PCA plots within each tissue and by family), My GLMMs will incorporate the family as random effect. By doing so, I aim to identify sets of genes, that are reliably differentially expressed in both families.


All models use following design: ~ Line + Conditioned + Experience + (1 | Family)
I'll use DESeq2 to acquire subsetted processed count data first

Following chunk filters the data and runs the GLMM for head tissue.


```{r GLMM head, echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE, collapse=TRUE}
# Subset the data for head tissue
#model for head
      # Subset the data for head tissue
      head_samples_data <- samples_data[samples_data$Tissue == "head", ]
      head_counts_data <- filtered_counts_data[, rownames(head_samples_data)]
      
      # Create DESeq2 object for head tissue
      dds_head <- DESeqDataSetFromMatrix(countData = head_counts_data,
                                        colData = head_samples_data,
                                        design = ~ Line + Conditioned + Experience + Family)
      
      # Perform DESeq normalization and dispersion estimation for head tissue
      dds_head <- DESeq(dds_head)
      dispersions_head <- dispersions(dds_head)
      dispersions_head <- setNames(dispersions_head, rownames(head_counts_data))
      size_factors_head <- sizeFactors(dds_head)
      
      # Normalize counts by size factors
      normalized_counts_head <- sweep(head_counts_data, 2, size_factors_head, "/")
      
      # Fit the glmmSeq model for head tissue
      #fit_head <- glmmSeq(~ Line + Conditioned + Experience + (1 | Family),
      #                    countdata = normalized_counts_head,
       #                   metadata = head_samples_data,
        #                  dispersion = dispersions_head,
         #                 family = "nbinom",
          #                progress = TRUE)
      
      # Save and print results for head tissue
      #saveRDS(fit_head, file = "fit_head_norm.rds")
      fit_head <- readRDS("fit_head_norm.rds")
          #fit_head <- readRDS("fit_head.rds") #alternatively to the counts normalized by size factors

      results_head <- summary(fit_head)
      #print(results_head)
      
      # Save the q-value results for head tissue
      qval_results_head <- glmmQvals(fit_head)
```      
The model for head tissue removed 2568 genes due to high singularity, which means that the coefficients' degree correlation is too high to calculate distinct effects from each separate coefficient.
However, for Line, the model highlighted 186, for Conditioned 904, and for Experience 38 significantly differentially expressed genes after correcting for family as random effect.

#### Model choice - Head
I'll also check other possible models (with interaction terms) and compare AIC scores in order to pick the best fitting model.


```{r GLMM head model choice, echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE, collapse=TRUE}

# Fit the remaining glmmSeq models for head tissue
#head2 <- glmmSeq(~ Line + Conditioned * Experience + (1 | Family),
 #                countdata = normalized_counts_head,
  #               metadata = head_samples_data,
   #              dispersion = dispersions_head,
    #             family = "nbinom",
     #            progress = TRUE)


                 
#head3 <- glmmSeq(~ Line * Conditioned * Experience + (1 | Family),
 #                countdata = normalized_counts_head,
  #               metadata = head_samples_data,
   #              dispersion = dispersions_head,
    #             family = "nbinom",
     #            progress = TRUE)



#head4 <- glmmSeq(~ Line * Conditioned + Experience + (1 | Family),
 #                countdata = normalized_counts_head,
  #               metadata = head_samples_data,
   #              dispersion = dispersions_head,
    #             family = "nbinom",
     #            progress = TRUE)



# Load your GLMM models (assuming they are saved as .RDS files in your working directory)

head1 <- readRDS("fit_head_norm.rds")
head2 <- readRDS("C:/Users/nanda/Desktop/Master Thesis/head2_glmmSeq.rds")
head3 <- readRDS("C:/Users/nanda/Desktop/Master Thesis/head3_glmmSeq.rds")
head4 <- readRDS("C:/Users/nanda/Desktop/Master Thesis/head4_glmmSeq.rds")

# Define a function to extract the correct AIC for GlmmSeq objects
extract_AIC_glmmSeq <- function(glmm_model) {
  # Access the statistics from the stats slot
  stats <- glmm_model@stats$res
  
  # Extract the AIC value directly from the stats if available
  if ("AIC" %in% colnames(stats)) {
    total_AIC <- sum(stats[, "AIC"], na.rm = TRUE)
  } else {
    # If AIC is not directly available, calculate it using log-likelihood
    logLik_values <- stats[, "logLik"]
    num_params <- ncol(glmm_model@stats$coef) + 1  # fixed effects + dispersion
    total_logLik <- sum(logLik_values, na.rm = TRUE)
    total_AIC <- -2 * total_logLik + 2 * num_params
  }
  
  return(total_AIC)
}

# Calculate AIC for each model using the correct method
AIC_head1 <- extract_AIC_glmmSeq(head1)
AIC_head2 <- extract_AIC_glmmSeq(head2)
AIC_head3 <- extract_AIC_glmmSeq(head3)
AIC_head4 <- extract_AIC_glmmSeq(head4)

# Print the AIC values
cat("AIC for head1:", AIC_head1, "\n")
cat("AIC for head2:", AIC_head2, "\n")
cat("AIC for head3:", AIC_head3, "\n")
cat("AIC for head4:", AIC_head4, "\n")
```
The initial model without interaction terms between the coefficients imply that every sub-treatment (line, conditioned, experience) operated independently from each other. For instance, Line can create evolutionary adaptations with fixed and robust expression patterns. The lack of interaction terms implies that plastic responses to emerging from a specific host type does not interact with traits that were potentially fixed over the generations of the line sub-treatment. This also applies to the unique single occasion of ovipositioning, which represents the experience sub-treatment.
The same pattern also applies to ovaries and venom glands.


#### Relative contributions of each sub-treatment to the differential gene expression - Head
Plot the significant genes in barplots, also create tables for relative contributions of treatment components with proportion of significant genes (not p-value!).


```{r GLMM head plots, echo=TRUE, fig.show='hold', collapse=TRUE}

#barplots for significant genes
     # Define a significance threshold
     q_threshold <- 0.05
                  
    # Extract the number of significant genes for each factor
    num_significant_genes <- c(
    Line = sum(qval_results_head@stats$qvals[, "Line"] < q_threshold),
    Conditioned = sum(qval_results_head@stats$qvals[, "Conditioned"] < q_threshold),
    Experience = sum(qval_results_head@stats$qvals[, "Experience"] < q_threshold))
                  
                  # Convert to data frame for plotting
                  significant_genes_df <- as.data.frame(num_significant_genes)
                  significant_genes_df$Factor <- rownames(significant_genes_df)
                  colnames(significant_genes_df) <- c("Count", "Factor")
                  
                  # Create the bar plot using ggplot2
                  library(ggplot2)
                  ggplot(significant_genes_df, aes(x = Factor, y = Count, fill = Factor)) +
                    geom_bar(stat = "identity") +
                    scale_fill_manual(values = c("#0072B2", "#E69F00", "#009E73")) +  # Colors: Blue, Orange, Green
                    labs(title = "Number of Significant Genes by Factor",
                         x = "Factor",
                         y = "Number of Significant Genes") +
                    theme_minimal()

                  
                  
#table + plots for coefficient of each factor + venn plots of overlapping genes
      # Extract coefficients and p-values for each factor by using correct indexing
      line_coefs <- results_head[, "LinePA"]
      conditioned_coefs <- results_head[, "ConditionedPA"]
      experience_coefs <- results_head[, "ExperiencePA"]
      
      line_pvals <- results_head[, "P_Line"]
      conditioned_pvals <- results_head[, "P_Conditioned"]
      experience_pvals <- results_head[, "P_Experience"]
      
      # Calculate summary statistics for each factor
      summary_stats <- data.frame(
        Factor = c("Line", "Conditioned", "Experience"),
        Mean_Coefficient = c(mean(line_coefs, na.rm = TRUE), 
                             mean(conditioned_coefs, na.rm = TRUE), 
                             mean(experience_coefs, na.rm = TRUE)),
        Median_Coefficient = c(median(line_coefs, na.rm = TRUE), 
                               median(conditioned_coefs, na.rm = TRUE), 
                               median(experience_coefs, na.rm = TRUE)),
        Proportion_Significant = c(mean(line_pvals < 0.05, na.rm = TRUE), 
                                   mean(conditioned_pvals < 0.05, na.rm = TRUE), 
                                   mean(experience_pvals < 0.05, na.rm = TRUE))
      )
      
      # Print summary statistics
      print(summary_stats)
      
      
      # Update the column names of summary_stats for better readability
      colnames(summary_stats) <- c("Factor", "Mean Coefficient", "Median Coefficient", "Proportion Significant")
      
      # Display the summary statistics using kable with updated column names
      kable(summary_stats, caption = "Summary of Factor Effects on Gene Expression") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
      
      
        # Combine data into a long format for plotting
      coefficients_df <- data.frame(
        Factor = rep(c("Line", "Conditioned", "Experience"), each = length(line_coefs)),
        Coefficient = c(line_coefs, conditioned_coefs, experience_coefs)
      )
      
      # Boxplot of coefficients to show the distribution of effects
      ggplot(coefficients_df, aes(x = Factor, y = Coefficient, fill = Factor)) +
        geom_boxplot() +
        labs(title = "Distribution of Coefficients by Factor", 
             y = "Coefficient Value", 
             x = "Factor") +
        theme_minimal()
      
      # Barplot of the proportion of significant genes for each factor
      ggplot(summary_stats, aes(x = Factor, y = `Proportion Significant`, fill = Factor)) +
        geom_bar(stat = "identity") +
        labs(title = "Proportion of Significant Genes by Factor", 
             y = "Proportion Significant (p < 0.05)", 
             x = "Factor") +
        theme_minimal()
      
#overlapping genes between sub-treatments
   
    # Load your GLMM results for head tissue
    fit_head <- readRDS("fit_head_norm.rds")
    
    # Extract q-values from the fit_head object
    qval_results_head <- glmmQvals(fit_head)
    
    # Define a significance threshold for q-values
    q_threshold <- 0.05
    
    # Extract significant genes for each factor in head tissue
    significant_genes_line_head <- rownames(qval_results_head@stats$qvals)[qval_results_head@stats$qvals[, "Line"] < q_threshold]
    significant_genes_conditioned_head <- rownames(qval_results_head@stats$qvals)[qval_results_head@stats$qvals[, "Conditioned"] < q_threshold]
    significant_genes_experience_head <- rownames(qval_results_head@stats$qvals)[qval_results_head@stats$qvals[, "Experience"] < q_threshold]
    
    # Calculate overlapping genes for each comparison in head tissue
    overlap_line_conditioned_head <- intersect(significant_genes_line_head, significant_genes_conditioned_head)
    overlap_line_experience_head <- intersect(significant_genes_line_head, significant_genes_experience_head)
    overlap_conditioned_experience_head <- intersect(significant_genes_conditioned_head, significant_genes_experience_head)
    
    # Calculate overlap between all three factors (Line, Conditioned, and Experience) for head tissue
    overlap_line_conditioned_experience_head <- Reduce(intersect, list(significant_genes_line_head, 
                                                                       significant_genes_conditioned_head, 
                                                                       significant_genes_experience_head))
    
    # Create a data frame for the number of overlapping genes in head tissue
    overlap_counts_df_head <- data.frame(
      Comparison = c("Line vs Conditioned", 
                     "Line vs Experience", 
                     "Conditioned vs Experience", 
                     "Line vs Conditioned vs Experience"),
      Overlapping_Genes = c(length(overlap_line_conditioned_head),
                            length(overlap_line_experience_head),
                            length(overlap_conditioned_experience_head),
                            length(overlap_line_conditioned_experience_head))
    )
    
    # Save the table of overlapping counts to a CSV file for head tissue
    write.csv(overlap_counts_df_head, file = "overlapping_gene_counts_head.csv", row.names = FALSE)
    
    
   # Rename the columns
    colnames(overlap_counts_df_head) <- c("Treatments", "Overlapping genes")

    # Display the table of overlapping counts in R (formatted for R Markdown)
    kable(overlap_counts_df_head, 
    caption = "Number of overlapping genes between Line, Conditioned, and Experience (Head)") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

```
Most significant genes from Experience (31 out of 38) overlap with the significant genes of Line and Conditioned. However, less than half of the significant genes of Line (186) overlap with the significant genes of Conditioned (904).

Conditioned genes show the two-fold number of significant genes (proportion significant) in comparison to Line. Experience has by the far the lowest number of significant genes.


### GLMM for ovaries
Ovaries


```{r GLMM ovaries, echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE, collapse=TRUE}
setwd("C:/Users/nanda/Desktop/Master Thesis")
 
# Subset the data for ovaries tissue
      ovaries_samples_data <- samples_data[samples_data$Tissue == "ovaries", ]
      ovaries_counts_data <- filtered_counts_data[, rownames(ovaries_samples_data)]
      
      # Create DESeq2 object for ovaries tissue
      dds_ovaries <- DESeqDataSetFromMatrix(countData = ovaries_counts_data,
                                            colData = ovaries_samples_data,
                                            design = ~ Line + Conditioned + Experience + Family)
      
      # Perform DESeq normalization and dispersion estimation for ovaries tissue
      dds_ovaries <- DESeq(dds_ovaries)
      dispersions_ovaries <- dispersions(dds_ovaries)
      dispersions_ovaries <- setNames(dispersions_ovaries, rownames(ovaries_counts_data))
      size_factors_ovaries <- sizeFactors(dds_ovaries)
      
      # Normalize counts by size factors
      normalized_counts_ovaries <- sweep(ovaries_counts_data, 2, size_factors_ovaries, "/")
      
      # Fit the glmmSeq model for ovaries tissue
      #fit_ovaries <- glmmSeq(~ Line + Conditioned + Experience + (1 | Family),
       #                      countdata = normalized_counts_ovaries,
        #                     metadata = ovaries_samples_data,
         #                    dispersion = dispersions_ovaries,
          #                   family = "nbinom",
           #                  progress = TRUE)
      
      # Save and print results for ovaries tissue
     #saveRDS(fit_ovaries, file = "fit_ovaries_norm.rds")
      fit_ovaries <- readRDS("fit_ovaries_norm.rds")
         #fit_ovaries <- readRDS("fit_ovaries.rds") #alternatively to the counts normalized by size factors
      
      results_ovaries <- summary(fit_ovaries)
      #print(results_ovaries)
      
      # Save the q-value results for ovaries tissue
      qval_results_ovaries <- glmmQvals(fit_ovaries)
      
```


#### Model choice - Ovaries
Check for best fitting model for ovaries too


```{r GLMM ovaries model choice, echo=TRUE, fig.show='hold', collapse=TRUE}
setwd("C:/Users/nanda/Desktop/Master Thesis")

# Model 2: Interaction between Conditioned and Experience
#ovaries2 <- glmmSeq(~ Line + Conditioned * Experience + (1 | Family),
 #                   countdata = normalized_counts_ovaries,
  #                  metadata = ovaries_samples_data,
   #                 dispersion = dispersions_ovaries,
    #                family = "nbinom",
     #               progress = TRUE)

# Model 3: Full interaction model
#ovaries3 <- glmmSeq(~ Line * Conditioned * Experience + (1 | Family),
 #                   countdata = normalized_counts_ovaries,
  #                  metadata = ovaries_samples_data,
   #                 dispersion = dispersions_ovaries,
    #                family = "nbinom",
     #               progress = TRUE)

# Model 4: Interaction between Line and Conditioned + Experience
#ovaries4 <- glmmSeq(~ Line * Conditioned + Experience + (1 | Family),
 #                   countdata = normalized_counts_ovaries,
  #                  metadata = ovaries_samples_data,
   #                 dispersion = dispersions_ovaries,
    #                family = "nbinom",
     #               progress = TRUE)

# Save models to disk for future use
#saveRDS(ovaries2, file = "ovaries2_glmmSeq.rds")
#saveRDS(ovaries3, file = "ovaries3_glmmSeq.rds")
#saveRDS(ovaries4, file = "ovaries4_glmmSeq.rds")

# Load your GLMM models for ovaries tissue (assuming these files are saved in your working directory)
ovaries1 <- readRDS("fit_ovaries_norm.rds")
ovaries2 <- readRDS("ovaries2_glmmSeq.rds")
ovaries3 <- readRDS("ovaries3_glmmSeq.rds")
ovaries4 <- readRDS("ovaries4_glmmSeq.rds")

# Define a function to extract AIC for GlmmSeq objects
extract_AIC_glmmSeq <- function(glmm_model) {
  # Access the statistics from the stats slot
  stats <- glmm_model@stats$res
  
  # Extract the AIC value directly from the stats if available
  if ("AIC" %in% colnames(stats)) {
    total_AIC <- sum(stats[, "AIC"], na.rm = TRUE)
  } else {
    # If AIC is not directly available, calculate it using log-likelihood
    logLik_values <- stats[, "logLik"]
    num_params <- ncol(glmm_model@stats$coef) + 1  # fixed effects + dispersion
    total_logLik <- sum(logLik_values, na.rm = TRUE)
    total_AIC <- -2 * total_logLik + 2 * num_params
  }
  
  return(total_AIC)
}

# Calculate AIC for each ovaries model
AIC_ovaries1 <- extract_AIC_glmmSeq(ovaries1)
AIC_ovaries2 <- extract_AIC_glmmSeq(ovaries2)
AIC_ovaries3 <- extract_AIC_glmmSeq(ovaries3)
AIC_ovaries4 <- extract_AIC_glmmSeq(ovaries4)

# Print the AIC values
cat("AIC for ovaries1:", AIC_ovaries1, "\n")
cat("AIC for ovaries2:", AIC_ovaries2, "\n")
cat("AIC for ovaries3:", AIC_ovaries3, "\n")
cat("AIC for ovaries4:", AIC_ovaries4, "\n")

```

Observations here are similar to those from the head model choice.


#### Relative contributions of each sub-treatment to the differential gene expression - Ovaries
Same procedure as for head tissue


```{r GLMM ovaries plots, echo=TRUE, fig.show='hold', collapse=TRUE}
#barplot significant genes
     # Define a significance threshold
     q_threshold <- 0.05
                  
     # Extract the number of significant genes for each factor in ovaries data
       num_significant_genes_ovaries <- c(
       Line = sum(qval_results_ovaries@stats$qvals[, "Line"] < q_threshold),
       Conditioned = sum(qval_results_ovaries@stats$qvals[, "Conditioned"] < q_threshold),
       Experience = sum(qval_results_ovaries@stats$qvals[, "Experience"] < q_threshold))
                  
                  # Convert to data frame for plotting
                  significant_genes_ovaries_df <- as.data.frame(num_significant_genes_ovaries)
                  significant_genes_ovaries_df$Factor <- rownames(significant_genes_ovaries_df)
                  colnames(significant_genes_ovaries_df) <- c("Count", "Factor")
                  
                  # Create the bar plot using ggplot2
                  library(ggplot2)
                  ggplot(significant_genes_ovaries_df, aes(x = Factor, y = Count, fill = Factor)) +
                    geom_bar(stat = "identity") +
                    scale_fill_manual(values = c("#0072B2", "#E69F00", "#009E73")) +  # Colors: Blue, Orange, Green
                    labs(title = "Number of Significant Genes by Factor (Ovaries)",
                         x = "Factor",
                         y = "Number of Significant Genes") +
                    theme_minimal()
                  
                  
#table + plots for coefficient of each factor
      # Extract coefficients and p-values for each factor by using correct indexing for ovaries tissue
      ovaries_line_coefs <- results_ovaries[, "LinePA"]
      ovaries_conditioned_coefs <- results_ovaries[, "ConditionedPA"]
      ovaries_experience_coefs <- results_ovaries[, "ExperiencePA"]
      
      ovaries_line_pvals <- results_ovaries[, "P_Line"]
      ovaries_conditioned_pvals <- results_ovaries[, "P_Conditioned"]
      ovaries_experience_pvals <- results_ovaries[, "P_Experience"]
      
      # Calculate summary statistics for each factor in ovaries tissue
      ovaries_summary_stats <- data.frame(
        Factor = c("Line", "Conditioned", "Experience"),
        Mean_Coefficient = c(mean(ovaries_line_coefs, na.rm = TRUE), 
                             mean(ovaries_conditioned_coefs, na.rm = TRUE), 
                             mean(ovaries_experience_coefs, na.rm = TRUE)),
        Median_Coefficient = c(median(ovaries_line_coefs, na.rm = TRUE), 
                               median(ovaries_conditioned_coefs, na.rm = TRUE), 
                               median(ovaries_experience_coefs, na.rm = TRUE)),
        Proportion_Significant = c(mean(ovaries_line_pvals < 0.05, na.rm = TRUE), 
                                   mean(ovaries_conditioned_pvals < 0.05, na.rm = TRUE), 
                                   mean(ovaries_experience_pvals < 0.05, na.rm = TRUE))
      )
      
      # Update the column names of ovaries_summary_stats for better readability
      colnames(ovaries_summary_stats) <- c("Factor", "Mean Coefficient", "Median Coefficient", "Proportion Significant")
      
      # Display the summary statistics for ovaries tissue using kable with updated column names
      kable(ovaries_summary_stats, caption = "Summary of Factor Effects on Gene Expression (Ovaries)") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
      
      # Combine data into a long format for plotting
      ovaries_coefficients_df <- data.frame(
        Factor = rep(c("Line", "Conditioned", "Experience"), each = length(ovaries_line_coefs)),
        Coefficient = c(ovaries_line_coefs, ovaries_conditioned_coefs, ovaries_experience_coefs)
      )
      
      # Boxplot of coefficients to show the distribution of effects for ovaries tissue
      ggplot(ovaries_coefficients_df, aes(x = Factor, y = Coefficient, fill = Factor)) +
        geom_boxplot() +
        labs(title = "Distribution of Coefficients by Factor (Ovaries)", 
             y = "Coefficient Value", 
             x = "Factor") +
        theme_minimal()
      
      # Barplot of the proportion of significant genes for each factor in ovaries tissue
      ggplot(ovaries_summary_stats, aes(x = Factor, y = `Proportion Significant`, fill = Factor)) +
        geom_bar(stat = "identity") +
        labs(title = "Proportion of Significant Genes by Factor (Ovaries)", 
             y = "Proportion Significant (p < 0.05)", 
             x = "Factor") +
        theme_minimal()
      
#overlapping genes
            
        
    # Load your GLMM results for ovaries tissue
    fit_ovaries <- readRDS("fit_ovaries_norm.rds")
    
    # Extract q-values from the fit_ovaries object
    qval_results_ovaries <- glmmQvals(fit_ovaries)
    
    # Define a significance threshold for q-values
    q_threshold <- 0.05
    
    # Extract significant genes for each factor in ovaries
    significant_genes_line_ovaries <- rownames(qval_results_ovaries@stats$qvals)[qval_results_ovaries@stats$qvals[, "Line"] < q_threshold]
    significant_genes_conditioned_ovaries <- rownames(qval_results_ovaries@stats$qvals)[qval_results_ovaries@stats$qvals[, "Conditioned"] < q_threshold]
    significant_genes_experience_ovaries <- rownames(qval_results_ovaries@stats$qvals)[qval_results_ovaries@stats$qvals[, "Experience"] < q_threshold]
    
    # Calculate overlapping genes for each comparison in ovaries
    overlap_line_conditioned_ovaries <- intersect(significant_genes_line_ovaries, significant_genes_conditioned_ovaries)
    overlap_line_experience_ovaries <- intersect(significant_genes_line_ovaries, significant_genes_experience_ovaries)
    overlap_conditioned_experience_ovaries <- intersect(significant_genes_conditioned_ovaries, significant_genes_experience_ovaries)
    
    # Calculate overlap between all three factors (Line, Conditioned, and Experience)
    overlap_line_conditioned_experience <- Reduce(intersect, list(significant_genes_line_ovaries, 
                                                                  significant_genes_conditioned_ovaries, 
                                                                  significant_genes_experience_ovaries))
    
    # Create a data frame for the number of overlapping genes
    overlap_counts_df <- data.frame(
      Comparison = c("Line vs Conditioned", 
                     "Line vs Experience", 
                     "Conditioned vs Experience", 
                     "Line vs Conditioned vs Experience"),
      Overlapping_Genes = c(length(overlap_line_conditioned_ovaries),
                            length(overlap_line_experience_ovaries),
                            length(overlap_conditioned_experience_ovaries),
                            length(overlap_line_conditioned_experience))
    )
    
    # Save the table of overlapping counts to a CSV file
    write.csv(overlap_counts_df, file = "overlapping_genes_ovaries.csv", row.names = FALSE)
    
    
    
    # Display the table of overlapping counts in R (formatted for R Markdown)
    kable(overlap_counts_df, caption = "Number of Overlapping Genes Between Line, Conditioned, and Experience (Ovaries)") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

Most significant genes from Line (57 in total) and Experience (58 in total) overlap with each other (44 overlaps). The majority of Line and Conditioned genes also overlap with Conditioned: 46 for Line, 51 for Experience.

Similarly to head, Conditioned has by far the highest number of significant genes in proportion to the total number of genes analyzed by the model, Experience the lowest.
However, the ovaries show the lowest number of significant genes for each sub-treatment in comparison to the other tissue types, head, and venom glands. Line only has one more significant genes than Experience, and overall, the difference between Conditioned and the other two sub-treatments is low.


### GLMM for venom glands
Venom tissue


```{r GLMM venom, echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE, collapse=TRUE}
 # Subset the data for venom tissue
      venom_samples_data <- samples_data[samples_data$Tissue == "venom", ]
      venom_counts_data <- filtered_counts_data[, rownames(venom_samples_data)]
      
      # Create DESeq2 object for venom tissue
      dds_venom <- DESeqDataSetFromMatrix(countData = venom_counts_data,
                                          colData = venom_samples_data,
                                          design = ~ Line + Conditioned + Experience + Family)
      
      # Perform DESeq normalization and dispersion estimation for venom tissue
      dds_venom <- DESeq(dds_venom)
      dispersions_venom <- dispersions(dds_venom)
      dispersions_venom <- setNames(dispersions_venom, rownames(venom_counts_data))
      size_factors_venom <- sizeFactors(dds_venom)
      
      
      # Normalize counts by size factors
      normalized_counts_venom <- sweep(venom_counts_data, 2, size_factors_venom, "/")
      
      
      # Fit the glmmSeq model for venom tissue
      #fit_venom <- glmmSeq(~ Line + Conditioned + Experience + (1 | Family),
       #                    countdata = normalized_counts_venom,
        #                   metadata = venom_samples_data,
         #                  dispersion = dispersions_venom,
          #                 family = "nbinom",
           #                progress = TRUE)
      
      # Save and print results for venom tissue
      #saveRDS(fit_venom, file = "fit_venom_norm.rds")
      fit_venom <- readRDS("fit_venom_norm.rds")
        #fit_venom <- readRDS("fit_venom.rds") #alternatively to the counts normalized by size factors
      
      results_venom <- summary(fit_venom)
      #print(results_venom)
      
      # Save the q-value results for venom tissue
      qval_results_venom <- glmmQvals(fit_venom)
```

#### model choice - venom glands
Same procedure and observation as in head and ovaries.

```{r GLMM venom model choice, eval=FALSE, fig.show='hold', collapse=TRUE, include=FALSE}
setwd("C:/Users/nanda/Desktop/Master Thesis")

# Model 2: Interaction between Conditioned and Experience
#venom2 <- glmmSeq(~ Line + Conditioned * Experience + (1 | Family),
 #                 countdata = normalized_counts_venom,
  #                metadata = venom_samples_data,
   #               dispersion = dispersions_venom,
    #              family = "nbinom",
     #             progress = TRUE)

# Model 3: Full interaction model
#venom3 <- glmmSeq(~ Line * Conditioned * Experience + (1 | Family),
 #                 countdata = normalized_counts_venom,
  #                metadata = venom_samples_data,
   #               dispersion = dispersions_venom,
    #              family = "nbinom",
     #             progress = TRUE)

# Model 4: Interaction between Line and Conditioned + Experience
#venom4 <- glmmSeq(~ Line * Conditioned + Experience + (1 | Family),
 #                 countdata = normalized_counts_venom,
  #                metadata = venom_samples_data,
   #               dispersion = dispersions_venom,
    #              family = "nbinom",
     #             progress = TRUE)

# Save models to disk for future use
#saveRDS(venom2, file = "venom2_glmmSeq.rds")
#saveRDS(venom3, file = "venom3_glmmSeq.rds")
#saveRDS(venom4, file = "venom4_glmmSeq.rds")


# Load your GLMM models for ovaries tissue (assuming these files are saved in your working directory)
venom1 <- readRDS("fit_venom_norm.rds")
venom2 <- readRDS("venom2_glmmSeq.rds")
venom3 <- readRDS("venom3_glmmSeq.rds")
venom4 <- readRDS("venom4_glmmSeq.rds")


# Define a function to extract AIC for GlmmSeq objects
extract_AIC_glmmSeq <- function(glmm_model) {
  # Access the statistics from the stats slot
  stats <- glmm_model@stats$res
  
  # Extract the AIC value directly from the stats if available
  if ("AIC" %in% colnames(stats)) {
    total_AIC <- sum(stats[, "AIC"], na.rm = TRUE)
  } else {
    # If AIC is not directly available, calculate it using log-likelihood
    logLik_values <- stats[, "logLik"]
    num_params <- ncol(glmm_model@stats$coef) + 1  # fixed effects + dispersion
    total_logLik <- sum(logLik_values, na.rm = TRUE)
    total_AIC <- -2 * total_logLik + 2 * num_params
  }
  
  return(total_AIC)
}

# Calculate AIC for each venom model
AIC_venom1 <- extract_AIC_glmmSeq(venom1)
AIC_venom2 <- extract_AIC_glmmSeq(venom2)
AIC_venom3 <- extract_AIC_glmmSeq(venom3)
AIC_venom4 <- extract_AIC_glmmSeq(venom4)

# Print the AIC values
cat("AIC for venom1:", AIC_venom1, "\n")
cat("AIC for venom2:", AIC_venom2, "\n")
cat("AIC for venom3:", AIC_venom3, "\n")
cat("AIC for venom4:", AIC_venom4, "\n")

```
The files are missing on my PC for some reason, I'll calculate the models again at some point, but the results were similar to those from head and ovaries.


#### Relative contributions of each sub-treatment to the differential gene expression - Venom glands
Same procedure as for head and ovaries.


```{r GLMM venom plots, echo=TRUE, fig.show='hold', collapse=TRUE}

#barplots for significant genes
    # Define a significance threshold
     q_threshold <- 0.05
            
    # Extract the number of significant genes for each factor in venom data
      num_significant_genes_venom <- c(
       Line = sum(qval_results_venom@stats$qvals[, "Line"] < q_threshold),
        Conditioned = sum(qval_results_venom@stats$qvals[, "Conditioned"] < q_threshold),
        Experience = sum(qval_results_venom@stats$qvals[, "Experience"] < q_threshold)
            )
            
            # Convert to data frame for plotting
            significant_genes_venom_df <- as.data.frame(num_significant_genes_venom)
            significant_genes_venom_df$Factor <- rownames(significant_genes_venom_df)
            colnames(significant_genes_venom_df) <- c("Count", "Factor")
            
            # Create the bar plot using ggplot2
            library(ggplot2)
            ggplot(significant_genes_venom_df, aes(x = Factor, y = Count, fill = Factor)) +
              geom_bar(stat = "identity") +
              scale_fill_manual(values = c("#0072B2", "#E69F00", "#009E73")) +  # Colors: Blue, Orange, Green
              labs(title = "Number of Significant Genes by Factor (Venom)",
                   x = "Factor",
                   y = "Number of Significant Genes") +
              theme_minimal()
            
#table + plots for coefficient of each factor
    # Extract coefficients and p-values for each 
     #factor by using correct indexing for venom tissue
      venom_line_coefs <- results_venom[, "LinePA"]
      venom_conditioned_coefs <- results_venom[, "ConditionedPA"]
      venom_experience_coefs <- results_venom[, "ExperiencePA"]
      
      venom_line_pvals <- results_venom[, "P_Line"]
      venom_conditioned_pvals <- results_venom[, "P_Conditioned"]
      venom_experience_pvals <- results_venom[, "P_Experience"]
      
      # Calculate summary statistics for each factor in venom tissue
      venom_summary_stats <- data.frame(
        Factor = c("Line", "Conditioned", "Experience"),
        Mean_Coefficient = c(mean(venom_line_coefs, na.rm = TRUE), 
                             mean(venom_conditioned_coefs, na.rm = TRUE), 
                             mean(venom_experience_coefs, na.rm = TRUE)),
        Median_Coefficient = c(median(venom_line_coefs, na.rm = TRUE), 
                               median(venom_conditioned_coefs, na.rm = TRUE), 
                               median(venom_experience_coefs, na.rm = TRUE)),
        Proportion_Significant = c(mean(venom_line_pvals < 0.05, na.rm = TRUE), 
                                   mean(venom_conditioned_pvals < 0.05, na.rm = TRUE), 
                                   mean(venom_experience_pvals < 0.05, na.rm = TRUE))
      )
      
      # Update the column names of venom_summary_stats for better readability
      colnames(venom_summary_stats) <- c("Factor", "Mean Coefficient", "Median Coefficient", "Proportion Significant")
      
      # Display the summary statistics for venom tissue using kable with updated column names
      kable(venom_summary_stats, caption = "Summary of Factor Effects on Gene Expression (Venom)") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
      
      # Combine data into a long format for plotting
      venom_coefficients_df <- data.frame(
        Factor = rep(c("Line", "Conditioned", "Experience"), each = length(venom_line_coefs)),
        Coefficient = c(venom_line_coefs, venom_conditioned_coefs, venom_experience_coefs)
      )
      
      # Boxplot of coefficients to show the distribution of effects for venom tissue
      ggplot(venom_coefficients_df, aes(x = Factor, y = Coefficient, fill = Factor)) +
        geom_boxplot() +
        labs(title = "Distribution of Coefficients by Factor (Venom)", 
             y = "Coefficient Value", 
             x = "Factor") +
        theme_minimal()
      
      # Barplot of the proportion of significant genes for each factor in venom tissue
      ggplot(venom_summary_stats, aes(x = Factor, y = `Proportion Significant`, fill = Factor)) +
        geom_bar(stat = "identity") +
        labs(title = "Proportion of Significant Genes by Factor (Venom)", 
             y = "Proportion Significant (p < 0.05)", 
             x = "Factor") +
        theme_minimal()

#overlapping genes
      
    # Load your GLMM results for venom tissue
    fit_venom <- readRDS("fit_venom_norm.rds")
    
    # Extract q-values from the fit_venom object
    qval_results_venom <- glmmQvals(fit_venom)
    
    # Define a significance threshold for q-values
    q_threshold <- 0.05
    
    # Extract significant genes for each factor in venom tissue
    significant_genes_line_venom <- rownames(qval_results_venom@stats$qvals)[qval_results_venom@stats$qvals[, "Line"] < q_threshold]
    significant_genes_conditioned_venom <- rownames(qval_results_venom@stats$qvals)[qval_results_venom@stats$qvals[, "Conditioned"] < q_threshold]
    significant_genes_experience_venom <- rownames(qval_results_venom@stats$qvals)[qval_results_venom@stats$qvals[, "Experience"] < q_threshold]
    
    # Calculate overlapping genes for each comparison in venom tissue
    overlap_line_conditioned_venom <- intersect(significant_genes_line_venom, significant_genes_conditioned_venom)
    overlap_line_experience_venom <- intersect(significant_genes_line_venom, significant_genes_experience_venom)
    overlap_conditioned_experience_venom <- intersect(significant_genes_conditioned_venom, significant_genes_experience_venom)
    
    # Calculate overlap between all three factors (Line, Conditioned, and Experience) for venom tissue
    overlap_line_conditioned_experience_venom <- Reduce(intersect, list(significant_genes_line_venom, 
                                                                        significant_genes_conditioned_venom, 
                                                                        significant_genes_experience_venom))
    
    # Create a data frame for the number of overlapping genes in venom tissue
    overlap_counts_df_venom <- data.frame(
      Comparison = c("Line vs Conditioned", 
                     "Line vs Experience", 
                     "Conditioned vs Experience", 
                     "Line vs Conditioned vs Experience"),
      Overlapping_Genes = c(length(overlap_line_conditioned_venom),
                            length(overlap_line_experience_venom),
                            length(overlap_conditioned_experience_venom),
                            length(overlap_line_conditioned_experience_venom))
    )
    
    # Save the table of overlapping counts to a CSV file for venom tissue
    write.csv(overlap_counts_df_venom, file = "overlapping_genes_venom.csv", row.names = FALSE)
    
    # Display the table of overlapping counts in R (formatted for R Markdown)
    kable(overlap_counts_df_venom, caption = "Number of Overlapping Genes Between Line, Conditioned, and Experience (Venom)") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

Again, the majority of significant genes of each sub-treatment overlap with each other. Also, the number of significant genes from experience is higher than in all remaining tissues, implying a higher relevance of the host encounter for venom gland transcriptome than for other tissues.

Similarly to head, Conditioned has by far the highest number of significant genes in proportion to the total number of genes analyzed by the model, Experience the lowest.

## Combined plots for GLMM

```{r Combined GLMM plots, echo=TRUE, fig.show='hold', collapse=TRUE}
# Libraries
library(ggplot2)
library(viridis)

# Example of significant gene counts across tissues
significant_genes_data <- data.frame(
  Tissue = rep(c("Head", "Ovaries", "Venom Glands"), each = 3),
  Sub_Treatment = rep(c("Line", "Conditioned", "Experience"), times = 3),
  Count = c(
    sum(qval_results_head@stats$qvals[, "Line"] < q_threshold),
    sum(qval_results_head@stats$qvals[, "Conditioned"] < q_threshold),
    sum(qval_results_head@stats$qvals[, "Experience"] < q_threshold),
    sum(qval_results_ovaries@stats$qvals[, "Line"] < q_threshold),
    sum(qval_results_ovaries@stats$qvals[, "Conditioned"] < q_threshold),
    sum(qval_results_ovaries@stats$qvals[, "Experience"] < q_threshold),
    sum(qval_results_venom@stats$qvals[, "Line"] < q_threshold),
    sum(qval_results_venom@stats$qvals[, "Conditioned"] < q_threshold),
    sum(qval_results_venom@stats$qvals[, "Experience"] < q_threshold)
  )
)

# Plot
ggplot(significant_genes_data, aes(x = Tissue, y = Count, fill = Sub_Treatment)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", linewidth = 0.8) +
  scale_fill_viridis_d(option = "C", direction = 1, name = "Sub-Treatment") +
  labs(
    title = "Number of Significant Genes by Sub-Treatments Across Tissues",
    x = NULL,
    y = "Number of Significant Genes"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.title.y = element_text(face = "bold", size = 14),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black", linewidth = 0.8),
    axis.line.y = element_line(color = "black", linewidth = 0.8),
    axis.ticks = element_line(color = "black")
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05))
  )

# Save the plot
ggsave(
  filename = "significant_genes_all_tissues.tiff",
  dpi = 300, width = 12, height = 6, units = "in", device = "tiff"
)


```


## Overlapping genes between tissue types

```{r GLMM tissue type overlap, echo=TRUE, fig.show='hold', collapse=TRUE}
# Load the GLMM results for each tissue
fit_head <- readRDS("fit_head_norm.rds")
fit_ovaries <- readRDS("fit_ovaries_norm.rds")
fit_venom <- readRDS("fit_venom_norm.rds")

# Extract q-values for each tissue
qval_results_head <- glmmQvals(fit_head)
qval_results_ovaries <- glmmQvals(fit_ovaries)
qval_results_venom <- glmmQvals(fit_venom)

# Define significance threshold
q_threshold <- 0.05

# Extract significant genes for each tissue type
significant_genes_head <- rownames(qval_results_head@stats$qvals)[qval_results_head@stats$qvals[, "Line"] < q_threshold]
significant_genes_ovaries <- rownames(qval_results_ovaries@stats$qvals)[qval_results_ovaries@stats$qvals[, "Line"] < q_threshold]
significant_genes_venom <- rownames(qval_results_venom@stats$qvals)[qval_results_venom@stats$qvals[, "Line"] < q_threshold]

# Calculate the pairwise overlaps between tissue types
overlap_head_ovaries <- intersect(significant_genes_head, significant_genes_ovaries)
overlap_head_venom <- intersect(significant_genes_head, significant_genes_venom)
overlap_ovaries_venom <- intersect(significant_genes_ovaries, significant_genes_venom)

# Calculate the overlap between all three tissue types
overlap_all_tissues <- Reduce(intersect, list(significant_genes_head, significant_genes_ovaries, significant_genes_venom))

# Create a data frame to show the overlap counts
overlap_counts_df <- data.frame(
  Comparison = c("Head vs Ovaries", "Head vs Venom", "Ovaries vs Venom", "All Tissues"),
  Overlapping_Genes = c(length(overlap_head_ovaries),
                        length(overlap_head_venom),
                        length(overlap_ovaries_venom),
                        length(overlap_all_tissues))
)

# Print the table in R Markdown using kable
kable(overlap_counts_df, caption = "Number of Overlapping Genes Between Tissue Types") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)

```

AS shown in the preliminary results from the PCA, there are no overlapping genes between all tissue types, and only few in pairwise comparisons of the tissue types.





## Volcano plots for significant genes from the models


```{r GLMM volcano plots, echo=TRUE, fig.show='hold', collapse=TRUE}

# Function to extract significant genes from glmmSeq results
extract_significant_genes <- function(glmm_model, factor_name, q_threshold = 0.05) {
  qvals <- glmmQvals(glmm_model)
  significant_genes <- rownames(qvals@stats$qvals)[qvals@stats$qvals[, factor_name] < q_threshold]
  return(significant_genes)
}

# Function to create volcano plots
create_volcano_plot <- function(glmm_model, factor_name, tissue_type) {
  results <- summary(glmm_model)
  df <- data.frame(
    Gene = rownames(results),
    logFC = results[, paste0(factor_name, "PA")],
    pvalue = results[, paste0("P_", factor_name)]
  )
  df$adj_pvalue <- p.adjust(df$pvalue, method = "fdr")
  df$Significant <- df$adj_pvalue < 0.05
  
  ggplot(df, aes(x = logFC, y = -log10(pvalue), color = Significant)) +
    geom_point(alpha = 0.8) +
    scale_color_manual(values = c("grey", "red")) +
    labs(title = paste("Volcano Plot for", factor_name, "in", tissue_type),
         x = "Log Fold Change",
         y = "-Log10(P-Value)") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Load GLMM models for each tissue type
# Assuming these files exist and have been saved as .RDS files
head1 <- readRDS("fit_head_norm.rds")
ovaries1 <- readRDS("fit_ovaries_norm.rds")
venom1 <- readRDS("fit_venom_norm.rds")

# Generate volcano plots for each tissue type and each factor (Line, Conditioned, Experience)

# Head Tissue
volcano_plot_head_line <- create_volcano_plot(head1, "Line", "Head Tissue")
volcano_plot_head_conditioned <- create_volcano_plot(head1, "Conditioned", "Head Tissue")
volcano_plot_head_experience <- create_volcano_plot(head1, "Experience", "Head Tissue")

# Ovaries Tissue
volcano_plot_ovaries_line <- create_volcano_plot(ovaries1, "Line", "Ovaries Tissue")
volcano_plot_ovaries_conditioned <- create_volcano_plot(ovaries1, "Conditioned", "Ovaries Tissue")
volcano_plot_ovaries_experience <- create_volcano_plot(ovaries1, "Experience", "Ovaries Tissue")

# Venom Tissue
volcano_plot_venom_line <- create_volcano_plot(venom1, "Line", "Venom Tissue")
volcano_plot_venom_conditioned <- create_volcano_plot(venom1, "Conditioned", "Venom Tissue")
volcano_plot_venom_experience <- create_volcano_plot(venom1, "Experience", "Venom Tissue")

# Display the plots
print(volcano_plot_head_line)
print(volcano_plot_head_conditioned)
print(volcano_plot_head_experience)

print(volcano_plot_ovaries_line)
print(volcano_plot_ovaries_conditioned)
print(volcano_plot_ovaries_experience)

print(volcano_plot_venom_line)
print(volcano_plot_venom_conditioned)
print(volcano_plot_venom_experience)


```


Check which side stands for PA/FA


```{r echo=TRUE, collapse = TRUE}

# If Line, Conditioned, and Experience are not factors, convert them to factors
samples_data$Line <- factor(samples_data$Line)
samples_data$Conditioned <- factor(samples_data$Conditioned)
samples_data$Experience <- factor(samples_data$Experience)

# Confirm the levels of each factor
levels(samples_data$Line)
levels(samples_data$Conditioned)
levels(samples_data$Experience)
```


Extract dataframes to save significant genes (q < 0.05) as csv.
The coefficient column represents the logfold change from the corresponding GLMM.
Genes with positive values are upregulated, genes with negative values downregulated.

The QValue column represents the adj. p-values. These values were adjusted for multiple testing by using 
the Benjamini-Hochberg method.


```{r echo=TRUE, collapse = TRUE}
#head
    # Extract q-values, coefficients, standard errors, and p-values from the GlmmSeq object
                #Line
                    # Obtain the q-values, coefficients, etc., for the head tissue
                    qvals_matrix_head <- qval_results_head@stats$qvals
                    coef_matrix_head <- qval_results_head@stats$coef
                    stderr_matrix_head <- qval_results_head@stats$stdErr
                    pvals_matrix_head <- qval_results_head@stats$pvals
                    
                    
                    # Set the significance threshold
                    q_threshold <- 0.05
                    
                    # For the Line factor (head tissue)
                    sig_genes_line_head <- rownames(qvals_matrix_head)[qvals_matrix_head[, "Line"] < q_threshold]
                    
                    # Create the data frame for Line using "LinePA" in coef_matrix and stderr_matrix
                    head_line_df <- data.frame(
                      Gene = sig_genes_line_head,
                      Coefficient = coef_matrix_head[sig_genes_line_head, "LinePA"],
                      StdError = stderr_matrix_head[sig_genes_line_head, "LinePA"],
                      PValue = pvals_matrix_head[sig_genes_line_head, "Line"],
                      QValue = qvals_matrix_head[sig_genes_line_head, "Line"]
                    )
                    
                #Conditioned
                    
                    # For the Conditioned factor (head tissue)
                    sig_genes_conditioned_head <- rownames(qvals_matrix_head)[qvals_matrix_head[, "Conditioned"] < q_threshold]
                    
                    # Create the data frame for Conditioned using "ConditionedPA" in coef_matrix and stderr_matrix
                    head_conditioned_df <- data.frame(
                      Gene = sig_genes_conditioned_head,
                      Coefficient = coef_matrix_head[sig_genes_conditioned_head, "ConditionedPA"],
                      StdError = stderr_matrix_head[sig_genes_conditioned_head, "ConditionedPA"],
                      PValue = pvals_matrix_head[sig_genes_conditioned_head, "Conditioned"],
                      QValue = qvals_matrix_head[sig_genes_conditioned_head, "Conditioned"]
                    )
              
                #Experience
                    
                    # For the Experience factor (head tissue)
                    sig_genes_experience_head <- rownames(qvals_matrix_head)[qvals_matrix_head[, "Experience"] < q_threshold]
                    
                    # Create the data frame for Experience using "ExperiencePA" in coef_matrix and stderr_matrix
                    head_experience_df <- data.frame(
                      Gene = sig_genes_experience_head,
                      Coefficient = coef_matrix_head[sig_genes_experience_head, "ExperiencePA"],
                      StdError = stderr_matrix_head[sig_genes_experience_head, "ExperiencePA"],
                      PValue = pvals_matrix_head[sig_genes_experience_head, "Experience"],
                      QValue = qvals_matrix_head[sig_genes_experience_head, "Experience"]
                    )
                    
              #save CSVs
                    write.csv(head_line_df, file = "head_line.csv", row.names = FALSE)
                    write.csv(head_conditioned_df, file = "head_conditioned.csv", row.names = FALSE)
                    write.csv(head_experience_df, file = "head_experience.csv", row.names = FALSE)
    
    
#ovaries
   # Extract q-values, coefficients, standard errors, and p-values from the GlmmSeq object
            # Obtain the q-values, coefficients, etc., for the ovaries tissue
            qval_results_ovaries <- glmmQvals(fit_ovaries)
            qvals_matrix_ovaries <- qval_results_ovaries@stats$qvals
            coef_matrix_ovaries <- qval_results_ovaries@stats$coef
            stderr_matrix_ovaries <- qval_results_ovaries@stats$stdErr
            pvals_matrix_ovaries <- qval_results_ovaries@stats$pvals
            
            # Set the significance threshold
            q_threshold <- 0.05
            
              #Line
                    sig_genes_line_ovaries <- rownames(qvals_matrix_ovaries)[qvals_matrix_ovaries[, "Line"] < q_threshold]
                    
                    # Create the data frame for Line using "LinePA" in coef_matrix and stderr_matrix
                    ovaries_line_df <- data.frame(
                      Gene = sig_genes_line_ovaries,
                      Coefficient = coef_matrix_ovaries[sig_genes_line_ovaries, "LinePA"],
                      StdError = stderr_matrix_ovaries[sig_genes_line_ovaries, "LinePA"],
                      PValue = pvals_matrix_ovaries[sig_genes_line_ovaries, "Line"],
                      QValue = qvals_matrix_ovaries[sig_genes_line_ovaries, "Line"]
                    )
                  
              #Conditioned
                    sig_genes_conditioned_ovaries <- rownames(qvals_matrix_ovaries)[qvals_matrix_ovaries[, "Conditioned"] < q_threshold]
                    
                    # Create the data frame for Conditioned using "ConditionedPA" in coef_matrix and stderr_matrix
                    ovaries_conditioned_df <- data.frame(
                      Gene = sig_genes_conditioned_ovaries,
                      Coefficient = coef_matrix_ovaries[sig_genes_conditioned_ovaries, "ConditionedPA"],
                      StdError = stderr_matrix_ovaries[sig_genes_conditioned_ovaries, "ConditionedPA"],
                      PValue = pvals_matrix_ovaries[sig_genes_conditioned_ovaries, "Conditioned"],
                      QValue = qvals_matrix_ovaries[sig_genes_conditioned_ovaries, "Conditioned"]
                    )
      
              #Experience
                    
                    # For the Experience factor (ovaries tissue)
                    sig_genes_experience_ovaries <- rownames(qvals_matrix_ovaries)[qvals_matrix_ovaries[, "Experience"] < q_threshold]
                    
                    # Create the data frame for Experience using "ExperiencePA" in coef_matrix and stderr_matrix
                    ovaries_experience_df <- data.frame(
                      Gene = sig_genes_experience_ovaries,
                      Coefficient = coef_matrix_ovaries[sig_genes_experience_ovaries, "ExperiencePA"],
                      StdError = stderr_matrix_ovaries[sig_genes_experience_ovaries, "ExperiencePA"],
                      PValue = pvals_matrix_ovaries[sig_genes_experience_ovaries, "Experience"],
                      QValue = qvals_matrix_ovaries[sig_genes_experience_ovaries, "Experience"]
                    )
              
              #save CSVs
                    write.csv(ovaries_line_df, file = "ovaries_line.csv", row.names = FALSE)
                    write.csv(ovaries_conditioned_df, file = "ovaries_conditioned.csv", row.names = FALSE)
                    write.csv(ovaries_experience_df, file = "ovaries_experience.csv", row.names = FALSE)                 

#venom
      # Extract q-values, coefficients, standard errors, and p-values from the GlmmSeq object
      qvals_matrix <- qval_results_venom@stats$qvals
      coef_matrix <- qval_results_venom@stats$coef
      stderr_matrix <- qval_results_venom@stats$stdErr
      pvals_matrix <- qval_results_venom@stats$pvals
      
      # Set the significance threshold
      q_threshold <- 0.05
      
      # For the Line factor
      sig_genes_line <- rownames(qvals_matrix)[qvals_matrix[, "Line"] < q_threshold]
      
            # Create the data frame for Line using "LinePA" in coef_matrix and stderr_matrix
              venom_line_df <- data.frame(
              Gene = sig_genes_line,
              Coefficient = coef_matrix[sig_genes_line, "LinePA"],
              StdError = stderr_matrix[sig_genes_line, "LinePA"],
              PValue = pvals_matrix[sig_genes_line, "Line"],
              QValue = qvals_matrix[sig_genes_line, "Line"]
            )
      
      
      # For the Conditioned factor
      sig_genes_conditioned <- rownames(qvals_matrix)[qvals_matrix[, "Conditioned"] < q_threshold]
            
            # Create the data frame for Conditioned using "ConditionedPA" in coef_matrix and stderr_matrix
              venom_conditioned_df <- data.frame(
              Gene = sig_genes_conditioned,
              Coefficient = coef_matrix[sig_genes_conditioned, "ConditionedPA"],
              StdError = stderr_matrix[sig_genes_conditioned, "ConditionedPA"],
              PValue = pvals_matrix[sig_genes_conditioned, "Conditioned"],
              QValue = qvals_matrix[sig_genes_conditioned, "Conditioned"]
            )
            
                  
      # For the Experience factor
      sig_genes_experience <- rownames(qvals_matrix)[qvals_matrix[, "Experience"] < q_threshold]
      
            # Create the data frame for Experience using "ExperiencePA" in coef_matrix and stderr_matrix
              venom_experience_df <- data.frame(
              Gene = sig_genes_experience,
              Coefficient = coef_matrix[sig_genes_experience, "ExperiencePA"],
              StdError = stderr_matrix[sig_genes_experience, "ExperiencePA"],
              PValue = pvals_matrix[sig_genes_experience, "Experience"],
              QValue = qvals_matrix[sig_genes_experience, "Experience"]
            )
              
              
          #save CSVs
                  #write.csv(venom_line_df, file = "venom_line.csv", row.names = FALSE)
                  #write.csv(venom_conditioned_df, file = "venom_conditioned.csv", row.names = FALSE)
                  #write.csv(venom_experience_df, file = "venom_experience.csv", row.names = FALSE)
```

   
## To do list
Remaining objectives:

- attempt to utilize the GLMM results to create heatmaps for treatment similarity based on the GLMM data.

- GO enrichment analysis with the sets of genes from the GLMMs DONE

- morpological quantification using phenopype DONE

- write Thesis and hand in